\documentclass[11pt]{article}

\input{../../preface/includes.tex}

\begin{document}

%    \setcounter{page}{1}
%    \begin{center}
%        {\Large {\bf Budget Justification}}
%    \end{center}

%    \begin{center}
%        {\Large Drake University}
%    \end{center}


\section*{Budget Justification}

	\subsection*{Senior Personnel (\$69,133)}
	PI - Dr. Titus Klinge will serve as the Principal Investigator (monthly salary \$8,222) and is requesting two months of salary each year of the project.  The year 1 amount is \$16,773, and an annual increase in salary is budgeted at 2\% per year.  Dr. Klinge will be responsible for leading the project at Drake University.
	
	\subsection*{Undergraduate Research Assistants (\$122,400)}
	Funding is requested for three undergraduate students each year at Drake University to work full-time for 10 weeks during the summer to participate in the project (\$72,000). In addition, funding is requested for three undergraduate students each year to work half-time during the academic year (\$50,400).

	\subsection*{Fringe Benefits (\$10,831)}
	Fringe benefits are requested at the current university rates of 7.7\% for faculty, and 7.65\% for undergraduate students.  Fringe benefits are charged to an individual based on his or her labor category.
	
	\subsection*{Travel (\$38,220)}
	The travel fund will be used to support the investigator and participating students to cover travel expenses associated with the collaboration between the three institutions:  Iowa State University, Grinnell College, and Drake University.  The travel fund will also be used to cover travel expenses for reporting results of this research at major national and international conferences.

\subsection*{Domestic}
\begin{center}
	\begin{tabular}{|p{4in}|}
		\hline 4 people (PI and students) -- Conference year 2, 3, and 4 \\ (e.g., DNA, POPL, SPLASH, PLDI) \\
		\hline Airfare \$450 per person \\
		\hline Hotel/Per Diem\\  4 nights = \$800 per person \\
		\hline Conference Registration \$450 per person \\
		\hline Total Conference years 2, 3, and 4 \$6,800 per year \$20,400 \\
		\hline \\
		\hline Collaboration travel to Drake University and Grinnell College  (Mileage 90 miles round trip (\$52) and meals two people \$28 ) \\
		\hline 6 trips per year at \$80 per trip = \$480  per year\\
		\hline Total collaboration travel for all years \$1920\\
		\hline \\
		\hline Total Domestic Travel \$22,320 \\
		\hline
	\end{tabular}
\end{center}

\subsection*{Foreign}
\begin{center}
	\begin{tabular}{|p{4in}|}
		\hline 2 people (PI and student) -- Conference year 2, 3, and 4 \\ (e.g., DNA, ICFP, SPLASH) \\
		\hline Airfare \$1,400 per person\\
		\hline Hotel/Per Diem\\  4 nights = \$800 per person \\
		\hline Conference Registration \$450 per person \\
		\hline \\
		\hline Total Foreign Travel \$15,900 \\
		\hline
	\end{tabular}
\end{center}
	
	\subsection*{Indirect Costs (\$88,294)}
	Drake University's DHHS negotiated on-campus indirect cost rate is 36.7\% of total direct costs excluding equipment, tuition allowance, patient care costs, and subcontract amounts exceeding \$25,000.


	\subsection*{}

\begin{tabular}{ll}
	{\large \bf Total Direct Costs:} & {\large \bf \$240,584 }\\
	{\large \bf Total Indirect Costs:} & {\large \bf \$88,294}\\
	{\large \bf Total Project Cost:} & {\large \bf \$328,878}
\end{tabular}

\end{document}