\documentclass[11pt]{article}

\newcommand{\thedoc}{Budget Justification}

\usepackage{proposal}

\begin{document}

\smalltitle{}

\subsection*{A. Senior Personnel: PI Peter-Michael Osera}
Senior personnel salary is requested for 2 months per summer in each year of the award.

\subsection*{B. Other Personnel}
\begin{itemize}[itemsep=0pt]
  \item \emph{Undergraduate Students (Summer Research)}: Stipend support is requested for \emph{four} Grinnell College undergraduate students conducting summer research in each year of the award (\$3400 per student).
    \$3400 is the college-standard stipend given to summer undergraduate research students.
    Students will work on projects related to the primary aims of this proposal in collaboration with the PI.
  \item \emph{Undergraduate Students (Hourly Workers)}: Hourly wage support is requested for two Grinnell College undergraduate students in each year of the award (\$1008 per student per semester).
    Hourly workers students will be expected to work 8 hours a week over a 14 week semester at the college-standard rate of \$9/hour.
    They will be tasked with data analysis and maintenance of software artifacts created during the course of this project.
\end{itemize}

\subsection*{C. Fringe Benefits}
\begin{itemize}[itemsep=0ex]
  \item The fringe rate for the PI is 17.65\%.
  \item The fringe rate for summer undergraduate student hourly workers (not including summer research students) and post-baccalaureate positions is 7.65\%.
\end{itemize}

\subsection*{E. Travel}

\paragraph{E.1 Domestic Travel}
Travel funding for the PI and four undergraduate students is requested to present preliminary work and network at domestic programming languages events, \eg, Midwest PL Summit (MWPLS) (\( \$1000 + 4 × \$1000 = \$5000 \) per year) as well as travel to meet with collaborators at Drake University and Iowa State University.
Undergraduates will alternate years presenting at regional and international venues (see E.2).

\paragraph{E.2 International Travel}
Travel is requested for one international conference per year for the PI and four undergraduate students (\( \$2500 + 4 × \$2500 = \$12500 \)) to present their work and network with the larger computer science research community.
Likely conferences include programming languages venues such as Programming Language Design and Implementation (PLDI), The International Conference on Functional Programming (ICFP), and Systems, Programming, Languages, and Applications: Software for Humanity (SPLASH), natural computing and molecular programming venues such as The International Conference on DNA Computing and Molecular Programming (DNA), and their associated workshops.

Undergraduates will alternate years presenting at regional and international venues (see E.1).

\subsection*{G. Other Direct Costs}

\subsection*{G.1. Equipment}
In the first year of this award, the PI will purchase two developer workstations (\eg, 4--6 core CPU, 32GB ram) and related peripherals (\$1250 per machine) to support their anticipated summer research students in their work.

\paragraph{G.2 Publication Costs/Documentation/Dissemination}
Funds are requested to provide open access for a published conference paper on the proposed work per year (\$900 per paper).

\paragraph{G.4 Computer Services}
Funds are requested to pay for monthly access to online services such as cloud servers through Linode for supporting development and hosting completed work (\$40 per month).

\subsection*{H. Indirect Costs}
The current negotiated overhead rate for Grinnell College is 65\% and is calculated on salaries and wages only.

\end{document}
