\documentclass[11pt]{article}

\input{../preface/includes.tex}

\begin{document}

    \setcounter{page}{1}
    \begin{center}
        {\Large {\bf Collaboration Plan}}
    \end{center}

      The PIs Klinge, Lathrop, and Osera are assistant professors at Drake University, Iowa State University, and Grinnell College, respectively.
      These institutions are all located within a 50 mile radius of each other.
      These three individuals will assume overall supervision of the proposed research.
      PI Lathrop and PI Klinge have relevant expertise in the theory of chemical reaction networks, PI Lathrop has some expertise in the software engineering of chemical reaction networks, and also industry experience in the engineering of large systems.
      PI Osera has relevant expertise in the theory of programming languages.
    
      Throughout the course of the proposed research PI Klinge will supervise 3 undergraduate students and PI Osera will supervise 2 undergraduate students each year at their respective institutions.
      PI~Lathrop and PI Klinge will co-supervise Ph.D. students from Iowa State University to help with the proposed research and mentoring of undergraduate students.
      In addition, two undergraduate research assistants from Iowa State University will participate in the proposed research during summers that the project is active.
      
      Three levels of communication will help facilitate the coordination of the PIs and students between the three institutions during the course of this proposed research.
      \begin{enumerate}
        \item
        Monthly on-site meetings that include formal presentations on the progress of the project.
        Time will also be allocated for in-depth technical discussions that are best suited for in-person meetings.
        These monthly meetings will also help foster a sense of community among the diverse participants from the three institutions.

        \item
        Weekly videoconference meetings between all participants that include short progress reports from each member.
        These meetings will give each participant prompt feedback on their progress from the other members of the project.
        This is especially important because of the interdisciplinary nature of the project and to keep the project moving forward.

        \item
        Real time persistent communication during the week via collaborative messaging tools such as Microsoft Teams, Slack, and collaborative code editing tools such as Microsoft Visual Studio Code's LiveShare extension.
        Tools such as Microsoft Teams and Slack preserve all communications that can easily be searched and archived.
      \end{enumerate}
      
      The two graduate research assistants will maintain elevated communication levels with undergraduate students at the three institutions.
      This will help keep undergraduate students working on the proposed research focused and any questions or issues addressed quickly.
      This also gives the graduate assistants experience in mentoring students under the supervision of PI Lathrop and PI Klinge.
      % Monthly more formal research presentations in person on location.
      % Weekly vitual meetings
      % Use of slack, visual code etc. to address real-time questions.
      
       
      Extensive use of audio/video conferencing and text based messaging services will be used by participants working on the project to communicate with other participants.
      It is anticipated that this informal communication medium will enhance the frequency and duration of interactions between collaborators at different institutions.
      Utilizing audio conference software, a weekly meeting will be held between all the participants in the proposed research.
      Also, the budget from each institution identifies funds for meetings at the other institutions' locations.
      The Iowa State University team plans to visit Drake University and Grinnell College each 3 times per year.
      Due to the teaching load and class schedules of students at undergraduate institutions, a reduced number of off-site trips are proposed for these institutions.
      These meetings will last most of a day and provide valuable face-to-face time for the participants in the proposed research to address any difficult or critical issues and present new ideas and results to others.
    
      The proposed research will utilize software servers and mirrored repositories at each of the institutions to share collaborative software, research artifacts, and publications.
    
\end{document}