%!TEX root = ../biographical_sketch.tex
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% PUBLICATIONS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\vspace*{-2ex}
\section*{Synergistic Activities}
\vspace*{-1ex}
\begin{enumerate}
    \item The PI developed an undergraduate course, \emph{Molecular Programming and Nanoscale Self-Assembly} surveying the results of twelve prominent research papers on molecular programming in the last decade.
    He taught the course at both Grinnell College and Carleton College and plans to offer a similar course at Drake University in the 2020/2021 academic year.
    \item Two students from the PI's winter 2019 molecular programming course at Carleton College continued their final projects as an independent study with the PI during the following spring term.
    These students, along with two graduate student collaborators from Iowa State University mentored by PI James Lathrop, presented their work as a poster at the 25th International Conference on DNA Computing and Molecular Programming in Seattle, WA in August 2019.
    Both Carleton students are now applying for graduate schools.
    \item The PI received a competitive \emph{mentored advanced project} (\emph{MAP}) grant from Grinnell College to hire four undergraduate students over the summer of 2017.
    These students researched various design patterns of molecular programming and presented their work at the Iowa Section of the Mathematical Association of America meeting in 2017.
    PI James Lathrop helped co-mentor these students from Iowa State University.
    The following summer, the PI co-mentored two undergraduate students at Iowa State University, continuing the project.
    \item The PI, along with PI James Lathrop, Jack Lutz, and Robyn Lutz, presented a full-day tutorial at the 30th International Conference on Automated Software Engineering in 2015 titled \emph{Molecular Programming: How Software Engineers Will Lead in the Age of Nanotechnology}.
    \item The PI, along with PI James Lathrop, organized a half-day undergraduate workshop at Simpson College in 2015 titled \emph{Molecular Programming}.

\end{enumerate}

% \section*{Refereed Conference Publications}
% \vspace{-2ex}
% \begin{longtable}{p{0.5in}|p{5.5in}}
%   2019 & \bibentry{rtcrn2} \\[1.5ex]
%   2018 & \bibentry{csc151} \\[1.5ex]
%   2017 & \bibentry{Ellis2017} \\[1.5ex]
%        & \bibentry{Huang2017} \\[1.5ex]
%   2016 & \bibentry{cKlinge16} \\[1.5ex]
%   2014 & \bibentry{cEHKLLL14}. \\
%        & \textbf{Manfred Paul Award for Excellence in Software: Theory and Practice}\\[1.5ex]
%   2012 & \bibentry{cLLLKMS12} \\[1.5ex]
%        & \bibentry{cLLLKHM12}
% \end{longtable}

% \section*{Technical Reports}
% \vspace{-2ex}
% \begin{longtable}{p{0.5in}|p{5.5in}}
%   2015 & \bibentry{oKlLaLu15}
% \end{longtable}