\documentclass[11pt]{article}

\input{../../preface/includes.tex}
\begin{document}

    \setcounter{page}{1}
    \begin{center}
        {\Large {\bf Facilities, Equipment, and Other Resources}}
    \end{center}

    \begin{center}
        {\Large Iowa State University}
    \end{center}

    \subsection*{Computer Science Computing Facilities}
    The Department of Computer Science at Iowa State University provides space and equipment for
    faculty members who are involved in teaching, graduate instruction,
    and research.   Technical and clerical support is also provided for
    the instructional and research activities for faculty and graduate
    students. The Computer Science Department has dedicated IT support
    from the system support group that reside in Atanasoff Hall.  The Computer Science group consists of two full time
    system analysts and an IT manager. Further field support is provided by
    student workers. The support group maintains 6 Windows laboratories
    with over 125 machines and a virtual Linux Laboratory. All students
    and faculty have access to these machines. 

    \subsection*{Liberal Arts and Sciences IT}
    As a part of the Liberal Arts and Sciences (LAS) IT organization, the systems support group
    works closely with employees who provide support across the LAS College. 
    The LAS  team supports several large clusters, research server grids,
    large scale research storage, virtual server and container based
    environments for research, web applications, and hundreds of
    scientific software applications.  The team is composed of a combined
    FTE of approximately five specialists who provide their expertise to
    consult on projects and manage the various services on offer. 

    \subsection*{Laboratory Facilities}
    Students in the PIs laboratories are provided with dockable laptops and/or workstations for their research and
    have access to ISU networks, storage, backup facilities and other computational resources. The University maintains
    shared storage (CyBox) and git facilities (GitLab) for all students
    and faculty.  The University has institutional licenses for common
    software required for this work such as MatLab. 

    PI Lathrop is part of the Laboratory for Molecular Programming (LAMP)
    laboratory which holds 14 students. The students working
    on this project will be housed in this laboratory.

    \subsection*{Office}
    The Computer Science Department and College of Liberal Arts and
    Sciences at Iowa State provides 
    clerical staff to support instructional and research activities, as
    well as for  travel, accounting support and grant management.

    \subsection*{Mentoring Resources}
    Although PI Lathrop will not receive summer salary as part of this proposal, he will be actively involved in the proposed research, and in the mentoring and supervision of graduate and undergraduate students involved in this proposed research.  In addition, PI Klinge is an Affiliate Assistant Professor in the Computer Science Department at Iowa State and will also co-mentor students in the department working on the proposed research.  PI Klinge has requested summer salary in part for this purpose.
\end{document}


