
\documentclass[11pt]{article}

\newcommand{\thedoc}{Budget Justification}
\usepackage{doc}

\begin{document}

\startdoc{\shorttitle}

\renewcommand{\thesection}{\Alph{section}}

\section{Senior Personnel: PI Peter-Michael Osera}

Senior personnel salary is requested for one summer month over each of the three years of the award.

\section{Other Personnel}

\begin{itemize}
  \item Undergraduate Students (Summer Research): Stipend support (\$3400 per student) is requested for 2 Grinnell College undergraduates conducting summer research in each year of the award.
\end{itemize}

\section{Fringe Benefits}

\begin{itemize}[itemsep=0pt]
  \item The fringe rate for the PI is 17.65\%.
\end{itemize}

\section{Equipment}

No equipment costs.

\section{Travel}

\subsection{Domestic Travel}

Travel funding is requested for visiting collaborators at their home institutions (Drake University and Iowa State University) up to six times a year as well as attending regional programming languages events, \eg, Midwest PL Summit (MWPLS), per year for the PI (\$1000).

Travel funding for two undergraduate students is requested for one year to present preliminary research at a regional programming languages event and interact with the programming languages community (\$1000 per student per year).

\subsection{International Travel}

Travel is requested for one international conference per year for the PI (\$2500).
Likely international conferences include international programming language conferences where the PI and his collaborators will present their work---POPL, PLDI, ICFP, and SPLASH---as well as international conferences in molecular programming.

International travel funding for two undergraduate students is requested for two years to present research at an international programming languages venue (\eg, the conferences listed above) and interact with the programming languages research community (\$2000 per student per year).

\section{Participant Support Costs}

No participant costs.

\section{Other Direct Costs}

No other direct costs.

\section{Indirect Costs}

The current negotiated overhead rate for Grinnell College is 62\% and is calculated on salaries and wages only.

\end{document}
