\documentclass[11pt]{article}

\input{../../preface/includes.tex}

\begin{document}

    \setcounter{page}{1}
    \begin{center}
        {\Large {\bf RUI Impact Statement}}
    \end{center}

    \begin{center}
        {\Large Drake University}
    \end{center}

    % %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \section{Drake University Profile}
    % %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    Located in Des Moines, IA, Drake University enrolls more than 3000 undergraduate students and more than 1800 masters students
    and is divided into seven colleges and schools.
    Drake University was founded in 1881 and is especially well-known for its Law School and Pharmacy and Health Sciences programs.
    The institution's largest school is the College of Arts and Sciences, consisting of over 47\% of student enrollment distributed over 42 majors.
    Drake attracts students from all over the world, with over 30 U.S.\ states and 28 countries represented.
    The College of Arts and Sciences maintains an 8:1 student-faculty ratio, and over 70\% of classes have fewer than 20 students enrolled.
    The faculty at Drake are passionate, and the small class sizes enhance pedagogy and active learning.
    As a result, Drake is ranked by U.S.\ News \& World Report in the top twelve institutions for undergraduate teaching in the Midwest.
    Drake students are ambitious; roughly one in every two students in the College of Arts and Sciences pursue a double major, triple major, or two or more minors.

    \textbf{Computer Science.}
    The computer science program at Drake University is still in its infancy but is rapidly developing; the number of majors has doubled every three years for over a decade.
    In 2018, there were 136 computer science majors at Drake in contrast to a single major in 2006.
    The program is housed in the Department of Mathematics and Computer Science along with six other programs, including data analytics and artificial intelligence.
    Combined, computer science, data analytics, and artificial intelligence have over 200 majors.
    Drake University is generously supportive of the rapidly growing department.
    The number of computer science tenure-track faculty at Drake has doubled in the last five years, and we are hiring two more computer science tenure-track faculty to start fall 2021.
    Moreover, the department recently moved into the newly constructed Collier-Scripps Hall, which features cutting-edge technology and generous classroom space to support high-quality pedagogy and scholarship.
    The computer science faculty are passionate about undergraduate education and have recently received multiple awards from the university for outstanding teaching effectiveness.

    \textbf{Research Environment.}
    Drake University provides generous support for faculty and student research.
    For example, the Drake Undergraduate Science Collaborative Institute (DUSCI) offers many programs to broaden undergraduate research participation.
    DUSCI offers \$3000 fellowships to students to participate in full-time eight-week summer research projects, supervised by Drake faculty.
    For seventeen years, the university has also hosted the Drake University Conference on Undergraduate Research in the Sciences (DUCURS), a full-day conference in the Spring semester.
    In 2019, DUCURS hosted 8 oral presentations and 84 posters.
    The Drake Science Colloquium Series is another program hosted by DUSCI and features presentations from Drake faculty and other scholars on their current research.

    The Mathematics and Computer Science Department is also exploring methods to broaden and diversify undergraduate research participation in computer science.
    For example, the department recently created a peer-mentored research program with projects overseen by department faculty.
    This fall, there are four active projects involving more than 30 students, many of whom are women and students from underrepresented groups.
    The department has also regularly funded student research through internal and external grants, such as the Iowa Space Consortium.

    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \section{Impact of Project}
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    Drake University is the PI's first tenure-track position, and this project will lay the foundation of his undergraduate research laboratory.
    The PI plans to maintain a group of 3--6 active undergraduate research students in perpetuity, and this project provides the initial support needed to launch his research group.
    The PI also plans to support additional students through his negotiated startup funds and other competitive grants such as DUSCI fellowships.

    \textbf{Preparing Students for Advanced Degrees in STEM.}
    Although the department's student-led research projects have successfully broadened and diversified the number of students interested in research, the projects themselves often lack the depth and the faculty-mentoring necessary to prepare students for advanced degrees in STEM.
    A major goal of this project is to create a cohort of undergraduate students, train them in cutting-edge molecular programming concepts, and mentor them through the scientific research process.
    During the project, three students per summer at Drake will experience 10 weeks of full-time research and will continue part-time during the academic year.
    All of the students will be exposed to graduate-level research and actively collaborate with students from Iowa State University and Grinnell College, including a graduate student funded by the project at Iowa State University.
    Each of the students will be individually mentored by the PI and will be provided the depth of research experience necessary to prepare them for advanced degrees in STEM.

    The PI is well-equipped to do this mentoring and has a history of preparing students for graduate school.
    While he was a Visiting Assistant Professor at Carleton College, the PI mentored two students on an independent research project in collaboration with two graduate students from Iowa State University supervised by PI James Lathrop.
    The students presented their work as a poster in 2019 and a peer-reviewed paper in 2020 at the \emph{International Conference on DNA Computing and Molecular Programming}.
    One of these students is from an underrepresented group, and both plan to apply for graduate schools in computer science.
    During the summer of 2020, the PI, along with PI James Lathrop, mentored one student from Carleton College and one student from Iowa State University on a summer research project that resulted in poster presentation.
    A paper is being drafted of these results to be submitted for review early next year, and the Carleton student is also planning to pursue a graduate degree in computer science.

    \textbf{Broadening Participation in STEM Research.}
    Molecular programming is a magnet for talented students.
    For example, in September, 2019, the PI gave a guest lecture in the Junior/Senior Chemistry Seminar at Drake University titled \emph{Molecular Programming: an algorithmic approach to DNA nanotechnology} and was immediately approached by multiple students interested in undergraduate research opportunities.
    Similarly, when the PI sought applicants for four undergraduate research assistant positions for a summer project at Grinnell College in 2017, a diverse pool of 24 qualified students applied.

    The PI has experience attracting students to STEM fields through his \emph{Introduction to Computer Science} courses at Drake University, Carleton College, and Grinnell College.
    While teaching these courses, many students from underrepresented groups approached the PI about their plans to add a computer science major after taking his class.
    For example, during the fall 2020 semester, one of the PI's female students of color added a second major in computer science and requested him as her advisor.
    Her first major is biochemistry, and the interdisciplinary nature of molecular programming makes it especially attractive to her.

    To help broaden participation in molecular programming, the PI will teach a special topics course during the 2021/2022 academic year titled \emph{Molecular Programming and Nanoscale Self-Assembly} at Drake University.
    This course will survey more than a dozen current research papers in molecular programming and expose students to a hands-on research experience.


    \textbf{Impact of Project on PI's Career}
    The project will have a strong impact on the PI's career.
    The project will enable the PI to travel to conferences with his students, which is essential during the first few years of an academic career.
    Furthermore, the project provides two months of summer research support to the PI per year to maintain his research productivity while also teaching a 3-3 course load.

    The PI's research agenda concerns bridging the gap of the theory and practice of molecular programming.
    This project is a natural next step to his research, and due to the interdisciplinary nature of the project, the collaborations with PI James Lathrop and PI Peter-Michael Osera are essential.
    Furthermore, the undergraduate students involved in the project will provide the support necessary to complete the research.

    \section{Plan for Finding Qualified Undergraduate Participants}
    In order to attract a diverse pool of talented students, the PI will teach a special topics course at Drake University on molecular programming during the 2021/2022 academic year.
    The course will help spark interest in the project as well as train students in the fundamentals of molecular programming before the first full-time summer of the project.
    To attract the interest of students outside the course, the PI will also advertise the project in fall 2021 when the department launches its student-led research projects.
    When appropriate, the PI will also personally encourage students from underrepresented groups to apply who may lack the self-confidence to consider joining the project.

    The criteria used for selecting undergraduate research assistants will include: academic success, passion for research, and clear interest in the project.
    First- and second-year students that demonstrate clear potential will also be considered.
    If applicants have similar qualifications, priority will be given to students from underrepresented groups.

    To assess the effect on the students participating on the project, the PI will monitor their academic achievements throughout the project and track how many participates go on to pursue a career in STEM or an advanced degree.

\end{document}