\documentclass[12pt,letterpaper]{article}
\begin{document}
	\section*{Budget Justification}
	
	%\subsection*{Senior Personnel}
	
	\subsection*{Graduate Research Assistants (\$74,598)}
	Funding is requested to support one graduate student to participate in the proposed research at 50\% effort for 12 months in years one and two, and for 9 months in year 3 @ \$2,200/month.  
	The fringe benefit rate for graduate research assistants at Iowa State University is 5.9\% of the salary.  The year 1 stipend amount is \$2,200/mo, and annual increase in salary is budgeted at 3\% per year.
	The year 1 salary is \$26,400.
	
	\subsection*{Undergraduate Research Assistants (\$22,254)}
	Funding is requested to support two undergraduate students each year for 3 months during the summer to participate in the proposed research.  The fringe benefit rate for undergraduate students at Iowa State University is 0.6\% of the wages.  The year 1 expected salary for each student is \$3,600 (\$1,200/mo) for a combined total of \$7,200 in year 1 for both students.
	An annual increase in salary is budgeted at 3\% per year.
	
	\subsection*{Fringe Benefits (\$4,534)}
	Fringe benefits are requested at the current university rates of 5.9\% for graduate students and 0.6\% for undergraduate students.  Fringe benefits are charged to an individual based on his or her labor category.
	
	\subsection*{Travel (\$6000)}
	The travel fund will be used to support the investigators and participating students to cover travel expenses associated with the collaboration between the three institutions:  Iowa State University, Grinnell College, and Drake University.  The travel fund will also be used to cover travel expenses for reporting results of this research at major national and international conferences.
	
	\subsection*{Domestic}
	\begin{center}
		Domestic Travel per Year (Multiply by 3)
	\begin{tabular}{|p{3in}|}
		\hline PI or GRA -- Conference 1 \\ (e.g., DNA conference, ASE conference) \\
		\hline Airfare \$520 \\
		\hline Hotel/Per Diem\\ \$150 x 4 nights = \$600 \\
		\hline Conference Registration \$400 \\

		\hline \\
		\hline Travel to Drake University and Grinnell College  (Mileage 90 miles round trip (\$52) and meals two people \$28 ) \\
		\hline 6 trips at \$80 per trip = \$480 \\
		\hline \\
		\hline Total: \$2000\\
		\hline
		
	\end{tabular}
\end{center}
	
	%subsection*{Foreign}
	
	%\subsection*{Other Direct Costs (\$45,626)}
	
	\subsection*{Tuition (\$44,322)}
	Funding for tuition is requested for one Ph.D. student in each year of the proposed research.  Tuition is listed at the rates approved by Iowa State University.
	
	\subsection*{Indirect Costs (\$56,914)}
	Iowa State University's DHHS negotiated on-campus indirect cost rate is 53\% of total direct costs excluding equipment, tuition allowance, patient care costs, and subcontract amounts exceeding \$25,000.
	
	\subsection*{}
	
	\begin{tabular}{ll}
			{\large \bf Total Direct Costs:} & {\large \bf \$151,708 }\\
			{\large \bf Total Indirect Costs:} & {\large \bf \$56,914}\\
			{\large \bf Total Project Cost:} & {\large \bf \$208,622}
	\end{tabular}

	
\end{document}